### 2022-03-01 paper/localpsf.tex
draft of interpolation section

Let $x$ and $x'$ be points in $\Omega$ that are close to each other, and consider the following three approximations:
\begin{align}
	\phi_x(y) &\approx \phi_{x'}(y) \label{eq:translate1}\\
	\phi_x(y) &\approx \phi_{x'}(y-x+x') \label{eq:translate2}\\
	\phi_x(y) &\approx \phi_{x'}\left(y-\mu(x)+\mu(x')\right). \label{eq:translate3}
\end{align}
Approximations \eqref{eq:translate1} and \eqref{eq:translate2} are standard. Approximation \eqref{eq:translate3}, which we use,  is new. These approximations are illustrated in Figure FIG.

@@

As discussed in Section SECTION, we have
\begin{equation*}
	\Phi(y,x) = \phi_x(y) \approx \phi_{x'}\left(y-\mu(x)+\mu(x')\right)
\end{equation*}
if $x'$ is close to $x$. Hence, we approximate the kernel entry $\Phi(y,x)$ by preconditioned radial basis function interpolation of the following (point,value) pairs:
\begin{equation}
	\left(x_i, \phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right)\right).
\end{equation}
Here $x_i \in \Omega$ are sample points at which the impulse response $\phi_{x_i}$ is known, and $x$ is the point we seek to evaluate the interpolation at.
%. Specifically, we interpolate $(\text{point}, \text{value})$ pairs $(x_i, f_i)$ at the point $x$, where $x_i$ are sample points, and $f_i$ are given as follows:
%\begin{equation}
%\label{eq:phi_mu_shift32}
%	f_i := \phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right).
%\end{equation}
Interpolation using all sample points $x_i$ is too expensive, so we only use the $k$ nearest sample points to $x$. In numerical experiments we use $k=10$. The $k$ nearest sample points are found by querying a precomputed KD-tree [CITE] of all sample points. We exclude points $x_i$ for which $\phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right)$ is not defined. In other words, if $y - \mu(x) + \mu(x_i) \notin \Omega$, then the point $x_i$ is considered invalid, and not used in the interpolation, even it is close to $x$. We check whether $y - \mu(x) + \mu(x_i) \in \Omega$ by querying a precomputed axis aligned bounding box tree (AABB tree) [CITE] of the mesh cells used to discretize the problem.

Once the $k$ nearest (valid) sample points, $\{x_i\}_{i=1}^k$, are found, we approximate $\Phi(y,x)$ using preconditioned radial basis function interpolation of the following form:
\begin{equation}
	\Phi(y,x) \approx \sum_{i=1}^k w_i~ b\left(\|x-x_i\|\right).
\end{equation}
For radial basis functions, $b_i$, we use polyharmonic splines. That is, 
\begin{equation}
b_i(r) = r^d
\end{equation}
if $d$ is odd, and
\begin{equation}
	b_i(r) = \begin{cases}
		r^d \log r, & r \neq 0 \\
		0 & r = 0
	\end{cases}
\end{equation}
if $d$ is even.
The vector of weights, $w = (w_1, w_2, \dots, w_k)^T$, is found as the solution to the $k \times k$ linear system
\begin{equation}
	Bw = f,
\end{equation}
where the coefficient matrix $B \in \mathbb{R}^{k \times k}$ has entries 
\begin{equation*}
	B_{ij} = b\left(\|x_i - x_j\|\right),
\end{equation*}
and the right hand side vector $f \in \mathbb{R}^k$ has entries
\begin{equation}
	f_i = \frac{V(x)}{V(x_i)}\phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right).
\end{equation}
%The result of this interpolation is to approximate $\Phi(y,x)$ with a linear combination of the shifted impulse response values $\phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right)$. 
By including the factor $\frac{V(x)}{V(x_i)}$, we are implicitly interpolating the ``volume preconditioned'' function $\frac{1}{V(x)}\Phi(y,x)$, then rescaling the result by $V(x)$ to get an approximation of $\Phi(y,x)$. Volume preconditioning improves the approximation considerably when $V(x)$ varies widely in magnitude, as is commonly the case for many operators of practical interest. In the Hessian for the Stokes inverse problem, for example, we observe that $V(x)$ varies over six orders of magnitude.

The point $y - \mu(x) + \mu(x_i)$ is typically not a gridpoint of the mesh used to discretize the problem, even if $y$, $x$, and $x_i$ are gridpoints. Although it is typically recommended that one avoid evaluating finite element functions at arbitrary points [CITE], here we have no choice. We evaluate $\frac{1}{V(x_i)}\phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right)$ via the following process. First, we determine whether $y - \mu(x) + \mu(x_i) \in E_{x_i}$ by checking if
\begin{equation}
\label{eq:ellipsoid_check93}
	(y - \mu(x))^T \Sigma^{-1}(x) (y - \mu(x))  \le \tau^2.
\end{equation}
If \eqref{eq:ellipsoid_check93} does not hold, then $y - \mu(x) + \mu(x_i) \notin E_{x_i}$, so we assume $\phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right)=0$ for the purpose of interpolation. If \eqref{eq:ellipsoid_check93} does hold, then $y - \mu(x) + \mu(x_i) \in E_{x_i}$. In this case, we look up the batch index $b$ for the batch of sample points containing $x_i$, and evaluate
\begin{equation}
	\frac{1}{V(x_i)}\phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right) = \eta_b\left(y - \mu(x) + \mu(x_i)\right).
\end{equation}
To evaluate $\eta_b\left(y - \mu(x) + \mu(x_i)\right)$, we recall the mesh cell that $y - \mu(x) + \mu(x_i)$ resides in. This mesh cell was already computed as a side effect of querying the AABB tree of mesh cells when we checked whether $y - \mu(x) + \mu(x_i) \in \Omega$. We then evaluate the relevant finite element basis functions on the cell to evaluate $\eta_b\left(y - \mu(x) + \mu(x_i)\right)$.

If $\Aop$ is symmetric, impulse responses of $\Aop$ are also impulse responses of $\Aop^T$. Let $\{y_i\}_{i=1}^k$ be the $k$ nearest sample points to $y$, excluding points $y_i$ such that $x - \mu(y) + \mu(y_i) \notin \Omega$. We could form an alternative approximation of $\Phi(y,x)$ by interpolating the (point, value) pairs
\begin{equation}
\label{eq:phi_mu_shift_y}
\left(y_i,\phi_{y_i}\left(x - \mu(y) + \mu(y_i)\right)\right)
\end{equation}
at the point $y$, in the manner described above but with $x$ and $y$ now taking opposite roles. We improve the interpolation accuracy at negligible additional cost by combining both of these approximations. Specifically, we combine both approximations by placing function values 
\begin{equation*}
	\left(\widetilde{f}_x\right)_i := \frac{V(x)}{V(x_i)}\phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right), \qquad i=1,\dots,k,
\end{equation*}
at the points $x_i-x$, placing function values
\begin{equation*}
\left(\widetilde{f}_y\right)_i := \frac{V(y)}{V(y_i)}\phi_{y_i}\left(x - \mu(y) + \mu(y_i)\right), \qquad i=1,\dots,k,
\end{equation*}
at the points $y_i-y$, and interpolating the combination of all $2k$ of these points and values at zero.

In detail, define 
\begin{equation*}
\widetilde{x_i} := x_i - x \qquad \text{and} \qquad \widetilde{y_i} := y_i - y.
\end{equation*}
and let
\begin{equation*}
	\widetilde{B} = \begin{bmatrix}
		\widetilde{B}_{xx} & \widetilde{B}_{xy} \\
		\widetilde{B}_{yx} & \widetilde{B}_{xx}
	\end{bmatrix}, \qquad \widetilde{w} = \begin{bmatrix}\widetilde{w}_x \\ \widetilde{w}_y\end{bmatrix}, \qquad \widetilde{f} = \begin{bmatrix}
		\widetilde{f}_x \\ \widetilde{f}_y
	\end{bmatrix},
\end{equation*}
be the $2k \times 2k$ block matrix and $2k \times 1$ block vectors, respectively, with block entries of $\widetilde{B}$ given by
\begin{align*}
\left(\widetilde{B}_{xx}\right)_{ij} = b\left(\|\widetilde{x}_i - \widetilde{x}_j\|\right), &\qquad
\left(\widetilde{B}_{xy}\right)_{ij} = b\left(\|\widetilde{x}_i - \widetilde{y}_j\|\right), \\
\left(\widetilde{B}_{yx}\right)_{ij} = b\left(\|\widetilde{y}_i - \widetilde{x}_j\|\right), &\qquad \left(\widetilde{B}_{yy}\right)_{ij} = b\left(\|\widetilde{y}_i - \widetilde{y}_j\|\right),
\end{align*}
for $i,j=1,\dots,k$. We approximate $\Phi(y,x)$ as follows:
\begin{equation}
\Phi(y,x) \approx \sum_{i=1}^k \left(w_x\right)_i~ b\left(\|\widetilde{x}_i\|\right) + \sum_{i=1}^k \left(w_y\right)_i~ b\left(\|\widetilde{y}_i\|\right),
\end{equation}
where $\widetilde{w}$ is the solution of the following regularized least-squares problem,
\begin{equation*}
	\min_{\widetilde{w}} \frac{1}{2}\|\widetilde{B} \widetilde{w} - \widetilde{f}\|^2 + \frac{\alpha}{2}\|\widetilde{w}\|^2,
\end{equation*}
for a small regularization parameter $\alpha$. 
We solve the regularized least squares problem by QR factorization of the block $4k \times 2k$ matrix $\begin{bmatrix}\widetilde{B} & I\end{bmatrix}^T$. We regularize the problem because of numerical issues that arise if there are points $\widetilde{x}_i$ and $\widetilde{y}_j$ in the interpolation such that $\widetilde{x}_i \approx \widetilde{y}_j$, a situation that arises often. Indeed, situations where $\widetilde{x}_i = \widetilde{y}_j$ are inevitable when a regular grid is used to discretize the problem. In this case, $\widetilde{B}$ is (nearly) singular, so it is numerically ill-advised to solve the standard interpolation equation $\widetilde{B} \widetilde{w} = \widetilde{f}$.

>>>>

Let $x$ and $x'$ be points in $\Omega$ that are close to each other, and consider the following three approximations, ordered by increasing level of sophistication:
\begin{align}
	\phi_x(y) &\approx \phi_{x'}(y) \label{eq:translate1}\\
	\phi_x(y) &\approx \phi_{x'}(y-x+x') \label{eq:translate2}\\
	\phi_x(y) &\approx \phi_{x'}\left(y-\mu(x)+\mu(x')\right). \label{eq:translate3}
\end{align}
Approximations \eqref{eq:translate1} and \eqref{eq:translate2} are standard. Approximation \eqref{eq:translate3}, which we use,  is new. These approximations are illustrated in Figure FIG.

@@

Given a pair of points $(y,x) \subset \Omega \times \Omega$, we define the function
\begin{equation}
\label{eq:fxyxp}
	f_{y,x}(x') := \frac{V(x)}{V(x')}\phi_{x_i}\left(y - \mu(x) + \mu(x')\right).
\end{equation}
It follows from inspection that 
\begin{equation*}
	\Phi(y,x) = f_{y,x}(x),
\end{equation*}
and from the discussion in Section SEC, we expect that $\Phi(y,x) \approx f_{y,x}(x')$ when $x'$ is close to $x$. The factor $\frac{V(x)}{V(x_i)}$, which was not discussed in Section SEC, is a form of preconditioning that improves the quality of this approximation in situations where $V(x)$ varies rapidly.
%, as is commonly the case for many operators of practical interest. In the Hessian for the Stokes inverse problem, for example, we observe that $V(x)$ varies over six orders of magnitude.
%As discussed in Section SECTION, we make the approximation
%\begin{equation*}
%	\Phi(y,x) = \phi_x(y) \approx \phi_{x'}\left(y-\mu(x)+\mu(x')\right)
%\end{equation*}
%if $x'$ is close to $x$. Hence, w

We approximate $\Phi(y,x)$ by performing radial basis function interpolation of the function $f_{y,x}(x')$ in the variable $x'$. The data used in the interpolation are the following (point,value) pairs
\begin{equation}
	\left(x_i, f_{y,x}(x_i)\right), \qquad i=1,\dots,k,
\end{equation}
and $(x, f_{y,x}(x))$ is the (point,value) pair we approximate via interpolation.
Here $\{x_i\}_{i=1}^k \subset \Omega$ are the $k$ nearest sample points to $x$, excluding sample points for which $f_{y,x}(x_i)$ is not defined.
%a collection of sample points at which the impulse response $\phi_{x_i}$ is known.
%. Specifically, we interpolate $(\text{point}, \text{value})$ pairs $(x_i, f_i)$ at the point $x$, where $x_i$ are sample points, and $f_i$ are given as follows:
%\begin{equation}
%\label{eq:phi_mu_shift32}
%	f_i := \phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right).
%\end{equation}
%We must perform a new interpolation every time we evaluate $\Phi(y,x)$ at a new pair of points $(y,x)$, and constructing the H-matrix approximation of $\Phi$ requires evaluating $\Phi(y,x)$ at a large number of pairs of points $(y,x)$. 
%Interpolation using all sample points $x_i$ is too expensive, so we only use the $k$ nearest sample points to $x$, excluding sample points for which $f_{y,x}(x_i)$ is not defined. 
In numerical experiments we use $k=10$. The $k$ nearest sample points are found by querying a precomputed KD-tree [CITE] of all sample points.

From the definition of $f_{y,x}$ in \eqref{eq:fxyxp}, we see that $f_{y,x}(x_i)$ is only defined if $y - \mu(x) + \mu(x_i) \in \Omega$, because $\phi_x$ is only defined on $\Omega$. But it may happen that $y - \mu(x) + \mu(x_i) \notin \Omega$, even if $y,x,x_i \in \Omega$. In this case, the point $x_i$ is considered invalid, and not used in the interpolation, even if it is close to $x$. We check whether $y - \mu(x) + \mu(x_i) \in \Omega$ by querying a precomputed axis aligned bounding box tree (AABB tree) [CITE] of the mesh cells used to discretize the problem.

We approximate $\Phi(y,x)$ using radial basis function interpolation of the following form:
\begin{equation}
	\Phi(y,x) = f_{y,x}(x) \approx \sum_{i=1}^k w_i~ b\left(\|x-x_i\|\right),
\end{equation}
where $b_i$ are radial basis functions, and $w_i$ are weights. The vector of weights, $w = (w_1, w_2, \dots, w_k)^T$, is found as the solution to the $k \times k$ linear system
\begin{equation}
Bw = f,
\end{equation}
where the coefficient matrix $B \in \mathbb{R}^{k \times k}$ has entries 
\begin{equation*}
B_{ij} = b\left(\|x_i - x_j\|\right),
\end{equation*}
and the right hand side vector $f \in \mathbb{R}^k$ has entries $f_i = f_{y,x}(x_i)$, $i=1,\dots,k$. We use polyharmonic spline radial basis functions [CITE], which are defined as follows:
\begin{equation}
	b_i(r) := \begin{cases}
		r^d, & d=1,3,5,\dots \\
		r^d \log r & d=2,4,6,\dots
	\end{cases}
\end{equation}
for $r>0$, and $b_i(0):=0$.
%\begin{equation}
%	f_i = \frac{V(x)}{V(x_i)}\phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right).
%\end{equation}
%The result of this interpolation is to approximate $\Phi(y,x)$ with a linear combination of the shifted impulse response values $\phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right)$. 
%By including the factor $\frac{V(x)}{V(x_i)}$, we are implicitly interpolating the ``volume preconditioned'' function $\frac{1}{V(x)}\Phi(y,x)$, then rescaling the result by $V(x)$ to get an approximation of $\Phi(y,x)$. Volume preconditioning improves the approximation considerably when $V(x)$ varies widely in magnitude, as is commonly the case for many operators of practical interest. In the Hessian for the Stokes inverse problem, for example, we observe that $V(x)$ varies over six orders of magnitude.

%In order to compute $f_{y,x}(x_i)$, we must evaluate $\impulseresponse_{x_i}\left(y - \mu(x) + \mu(x_i)\right)$, and we do this by evaluating $\eta_b\left(y - \mu(x) + \mu(x_i)\right)$. Here $b$ is the index of the batch that contains the sample point $x_i$, and $\eta_b$ is the corresponding impulse response batch function that we have computed and stored.  
%Although it is typically recommended that one avoid evaluating finite element functions at arbitrary points [CITE], here we have no choice. 
We evaluate $f_{y,x}(x_i)$ via the following process. First, we determine whether $y - \mu(x) + \mu(x_i) \in E_{x_i}$ by checking if
\begin{equation}
\label{eq:ellipsoid_check93}
	(y - \mu(x))^T \Sigma^{-1}(x) (y - \mu(x))  \le \tau^2.
\end{equation}
If \eqref{eq:ellipsoid_check93} does not hold, then $y - \mu(x) + \mu(x_i) \notin E_{x_i}$ (i.e., $y - \mu(x) + \mu(x_i)$ is outside the estimated support of $\impulseresponse_{x_i}$), so we set $f_{y,x}(x_i)=0$ for the purpose of interpolation. If \eqref{eq:ellipsoid_check93} does hold, then $y - \mu(x) + \mu(x_i) \in E_{x_i}$. In this case, we look up the batch index $b$ for the batch of sample points that contains $x_i$, and evaluate
\begin{equation}
	f_{y,x}(x_i) = \frac{V(x)}{V(x_i)}\phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right) = V(x)\eta_b\left(y - \mu(x) + \mu(x_i)\right),
\end{equation}
where $\eta_b$ is the impulse response batch function that we have computed and stored.
Evaluating $\eta_b\left(y - \mu(x) + \mu(x_i)\right)$ requires some care, because the point $y - \mu(x) + \mu(x_i)$ is typically not a gridpoint of the mesh used to discretize the problem, even if $y$, $x$, and $x_i$ are gridpoints.
To evaluate $\eta_b\left(y - \mu(x) + \mu(x_i)\right)$, we determine which mesh cell contains $y - \mu(x) + \mu(x_i)$, then we evaluate the relevant finite element basis functions on that mesh cell. Fortunately, the mesh cell containing $y - \mu(x) + \mu(x_i)$ was already determined as a side effect of querying the AABB tree of mesh cells when we checked whether $y - \mu(x) + \mu(x_i) \in \Omega$.

If $\Aop$ is symmetric, impulse responses of $\Aop$ are also impulse responses of $\Aop^T$. Let $\{y_i\}_{i=1}^k$ be the $k$ nearest sample points to $y$, excluding points $y_i$ such that $x - \mu(y) + \mu(y_i) \notin \Omega$. We could form an alternative approximation of $\Phi(y,x)$ by interpolating the (point, value) pairs
\begin{equation}
\label{eq:phi_mu_shift_y}
\left(y_i,f_{x,y}(y_i)\right), \qquad i=1,\dots,k,
\end{equation}
in the same manner as before, but with $x$ and $y$ now taking opposite roles. We improve the interpolation accuracy at negligible additional cost by combining both of these approximations. Specifically, we combine both approximations by placing function values 
\begin{equation*}
	\left(\widetilde{f}_x\right)_i := \frac{V(x)}{V(x_i)}\phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right), \qquad i=1,\dots,k,
\end{equation*}
at the points $x_i-x$, placing function values
\begin{equation*}
\left(\widetilde{f}_y\right)_i := \frac{V(y)}{V(y_i)}\phi_{y_i}\left(x - \mu(y) + \mu(y_i)\right), \qquad i=1,\dots,k,
\end{equation*}
at the points $y_i-y$, and interpolating the combination of all $2k$ of these points and values at zero.

In detail, define 
\begin{equation*}
\widetilde{x_i} := x_i - x \qquad \text{and} \qquad \widetilde{y_i} := y_i - y.
\end{equation*}
and let
\begin{equation*}
	\widetilde{B} = \begin{bmatrix}
		\widetilde{B}_{xx} & \widetilde{B}_{xy} \\
		\widetilde{B}_{yx} & \widetilde{B}_{xx}
	\end{bmatrix}, \qquad \widetilde{w} = \begin{bmatrix}\widetilde{w}_x \\ \widetilde{w}_y\end{bmatrix}, \qquad \widetilde{f} = \begin{bmatrix}
		\widetilde{f}_x \\ \widetilde{f}_y
	\end{bmatrix},
\end{equation*}
be a $2k \times 2k$ block matrix and $2k \times 1$ block vectors, respectively, with the block entries of $\widetilde{B}$ be given by
\begin{align*}
\left(\widetilde{B}_{xx}\right)_{ij} = b\left(\|\widetilde{x}_i - \widetilde{x}_j\|\right), &\qquad
\left(\widetilde{B}_{xy}\right)_{ij} = b\left(\|\widetilde{x}_i - \widetilde{y}_j\|\right), \\
\left(\widetilde{B}_{yx}\right)_{ij} = b\left(\|\widetilde{y}_i - \widetilde{x}_j\|\right), &\qquad \left(\widetilde{B}_{yy}\right)_{ij} = b\left(\|\widetilde{y}_i - \widetilde{y}_j\|\right),
\end{align*}
for $i,j=1,\dots,k$. We approximate $\Phi(y,x)$ as follows:
\begin{equation}
\Phi(y,x) \approx \sum_{i=1}^k \left(w_x\right)_i~ b\left(\|\widetilde{x}_i\|\right) + \sum_{i=1}^k \left(w_y\right)_i~ b\left(\|\widetilde{y}_i\|\right),
\end{equation}
where $\widetilde{w}$ is the solution of the following regularized least-squares problem,
\begin{equation*}
	\min_{\widetilde{w}} \frac{1}{2}\|\widetilde{B} \widetilde{w} - \widetilde{f}\|^2 + \frac{\alpha}{2}\|\widetilde{w}\|^2,
\end{equation*}
for a small regularization parameter $\alpha$. 
We solve the regularized least squares problem by QR factorization of the block $4k \times 2k$ matrix $\begin{bmatrix}\widetilde{B} & I\end{bmatrix}^T$. We regularize the problem because of numerical issues that arise if there are points $\widetilde{x}_i$ and $\widetilde{y}_j$ in the interpolation such that $\widetilde{x}_i \approx \widetilde{y}_j$, a situation that arises often in practice. Indeed, situations where $\widetilde{x}_i = \widetilde{y}_j$ are inevitable when a regular grid is used to discretize the problem. In this case, $\widetilde{B}$ is (nearly) singular, so it is numerically ill-advised to solve the standard interpolation equation $\widetilde{B} \widetilde{w} = \widetilde{f}$.
